---
title: ""
author: ""
date: ""
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE, echo=TRUE}

# Global NA to blank
options(knitr.kable.NA = '')
knitr::opts_chunk$set(echo = FALSE)

# Libraries
library(sqldf)
library(tidyverse)
library(odbc)
library(leaflet)
library(psych)
library(GGally)
library(ggpubr)

# Database connection 
connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inpyugamsvm01\\nuna", Database = "AK_ShallowLakes")

# Figure and Table Counters
FigureCounter = 1
TableCounter = 1

```

```{r,label="Get water profile chemical dataset"}
# Get wate profile parameters by depth for the continuous water quality lake visits
Sql = " SELECT [PONDNAME]
,[SAMPLEDATE]
,Month(SampleDate) as [Month]
,Case When Month(SampleDate) < 7 Then 'Spring' Else 'Fall' End as Season
,[SAMPLEDEPTH]
,[Temperature_Mean]
,[Temperature_Min]
,[Temperature_Max]
,[Temperature_SD]
,[Temperature_n]
,[pH_Mean]
,[pH_Min]
,[pH_Max]
,[pH_SD]
,[pH_n]
,[SpCond_Mean]
,[SpCond_Min]
,[SpCond_Max]
,[SpCond_SD]
,[SpCond_n]
,[DO_Mean]
,[DO_Min]
,[DO_Max]
,[DO_SD]
,[DO_n]
,[DO_Pct_Mean]
,[DO_Pct_Min]
,[DO_Pct_Max]
,[DO_Pct_SD]
,[DO_Pct_n]
FROM [AK_ShallowLakes].[dbo].[Summary_WaterProfileStatisticsByLake]
ORDER BY PONDNAME,SAMPLEDATE,SAMPLEDEPTH"
Means = dbGetQuery(connection,Sql)

# Make a counter to keep track of the QC test numbers
QCNum = 1
```

```{r, label="Get the visits timeline dataset"}
# Get the continuous water quality visits history for the lakes
Sql = "SELECT  w.PONDNAME, YEAR(w.SAMPLEDATE) AS Year, COUNT(*) AS n
FROM   tblWaterProfiles AS w INNER JOIN
       tblEvents AS e ON w.PONDNAME = e.PONDNAME
WHERE  (e.CONTMONVISIT = 1)
GROUP BY w.PONDNAME, YEAR(w.SAMPLEDATE)
ORDER BY w.PONDNAME, YEAR(w.SAMPLEDATE)"
Visits = dbGetQuery(connection,Sql)

```

# Continuous data analysis, 2024

Objective: An analysis of the ProDss data at the 0.5 m depth across all the continuous monitoring lakes in CAKN and ARCN. Regress temperature, spcond and pH.

# Data Profiles By Lake

```{r, label="Visits summary",results='asis'}

GetReportSectionForLake = function(Lake){
  FigureCounter = 1
  cat("## ",Lake,"  \n\n")
  
  # Plot the visits history
  cat("### Site visit history and number of records collected  \n\n")
  print(
    ggplot(Visits %>% filter(PONDNAME==Lake)) +
      geom_col(aes(x=Year,y=n),na.rm=TRUE,width = 0.5) +
      ggtitle(paste(Lake," Site visit history and sampling intensity",sep="")) +
      theme_minimal() +
      ylab("Number of records") +
      theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
      scale_x_continuous(breaks = seq(min(Visits$Year), max(Visits$Year), by = 1))
    )
  cat("\n\nFigure ",Lake,"-",FigureCounter,". Lake visitation history, ",Lake,".  \n\n")
  FigureCounter = FigureCounter + 1
  
  
  # Plot a parameter vs depth plot for each parameter
  # Loop through the column names
  for (ColumnName in colnames(Means)) {
    
    # Isolate the parameters (they have '_Mean' suffixes)
    if(grepl("_Mean",ColumnName) == TRUE){
      
      # Get a more readable parameter name by removing the '_Mean' suffix and output it as a header
      ParameterName = gsub("_Mean","",ColumnName)
      cat("### ",ParameterName,"  \n\n")
      
      # Get the data for the lake and parameter
      Sql = paste("SELECT PONDNAME,SAMPLEDATE,SAMPLEDEPTH * -1 as [Depth (m)],Season
,",ColumnName,",",gsub("_Mean","_SD",ColumnName),",",gsub("_Mean","_Min",ColumnName),",",gsub("_Mean","_Max",ColumnName)," ,",gsub("_Mean","_n",ColumnName)," 
 FROM Means WHERE PONDNAME='",Lake,"' AND ",ColumnName," IS NOT NULL 
ORDER BY PONDNAME,SAMPLEDATE,SAMPLEDEPTH",sep="")
      data = sqldf(Sql)

      
      # Plot a histogram showing the distribution of the binned parameter data
      cat("#### Distribution of values  \n\n")
      # Generate some reasonable bin widths for plotting a histogram of the parameter data
      Binwidth <- (max(data[ColumnName],na.rm=TRUE) - min(data[ColumnName],na.rm=TRUE)) / ceiling(log2(nrow(data)) + 1)
      print(
        ggplot(data) + 
          geom_histogram(aes_string(x=ColumnName),binwidth=Binwidth,na.rm=TRUE) + 
          facet_wrap(vars(Season)) +
          ylab(paste("Count (",nrow(data)," non-null rows)",sep="")) + 
          theme_minimal()) +
          ggtitle(paste(Lake,": Distribution Of Values: ",ParameterName,sep="")
      ) 
      cat("\n\nFigure ",Lake,"-",FigureCounter,"Histogram of ",ParameterName,". Bin width = ",Binwidth,".  \n\n")
      FigureCounter = FigureCounter + 1
      
      
      
      # Calculate and summarize in a table basic statistics about the chemical component
      # cat("#### Basic statistics  \n\n")
      # cat("\n\nTable ",Lake,"-1. Basic statistics, ",ParameterName,".  \n\n")
      # Stats = describe(data[ColumnName])
      # Stats = Stats %>% select(mean,	sd,	min,	max,	range,	skew,	kurtosis,	se,n)
      # print(knitr::kable(t(Stats)))
      # cat("  \n\n")

      cat("#### Seasonal difference at 0.5m depth  \n\n")
      cat("'Season' is broadly defined as Month <=7 = Spring, Month > 7 = Fall.   \n\n")

      
      # Show a boxplot with t-test result
      print(
        ggplot(data) +
          geom_boxplot(aes_string(x = 'Season', y = ColumnName),na.rm=TRUE) +
          stat_compare_means(aes_string(x = 'Season', y = ColumnName),na.rm=TRUE,method = "t.test") +  # Add t-test comparison
          theme_minimal() +
          ggtitle(paste(Lake," Seasonal difference in ",ParameterName," at 0.5m depth with t-test significance level",sep="")) 
      )
      cat("\n\nFigure ",Lake,"-",FigureCounter,". Seasonal difference in ",ParameterName," at 0.5m depth with t-test significance level in lake ",Lake,".  \n\n")
      FigureCounter = FigureCounter + 1
      
      # Perform t-test
      t_test_result = tryCatch({
        t_test_result <- t.test(data[[ColumnName]] ~ Season, data = data)
        cat("**T-Test**  \n\n")
        # Print the result
        cat("**Statistic:**", t_test_result$statistic, "  \n")
        cat("**P-Value:**", t_test_result$p.value, "  \n")
        cat("**Confidence Interval:**", t_test_result$conf.int, "  \n")
        cat("**Means:**", t_test_result$estimate, "  \n\n")
        cat("  \n\n")
      }, error = function(e) {
        cat("Error: ", conditionMessage(e))
        return(NA)
      })

      # Plot the parameter for the lake by season
      cat("#### Depth profile  \n\n")
      cat("'Season' is broadly defined as Month <=7 = Spring, Month > 7 = Fall.   \n\n")
      print(
        ggplot(data) +
          geom_path(aes_string(x=ColumnName,y="`Depth (m)`",color='Season',linetype='SAMPLEDATE'),na.rm=TRUE,size = 1) +
          scale_color_manual(values = c("Spring" = "darkgreen", "Fall" = "brown")) +
          theme_minimal() +
          ggtitle(paste(Lake," Depth Profile by season: ",ParameterName,sep="")) 
      )
      cat("\n\nFigure ",Lake,"-",FigureCounter,"Depth profile by season, ",ParameterName,".  \n\n")
      FigureCounter = FigureCounter + 1
      
    }
  }
}

#  GetReportSectionForLake('YUCH-004')

# Lake='YUCH-004'
# ColumnName='DO_Mean'
# # Get the data for the lake and parameter
# Sql = paste("SELECT PONDNAME,SAMPLEDATE,SAMPLEDEPTH * -1 as [Depth (m)],Season
# ,",ColumnName,",",gsub("_Mean","_SD",ColumnName),",",gsub("_Mean","_Min",ColumnName),",",gsub("_Mean","_Max",ColumnName)," ,",gsub("_Mean","_n",ColumnName),"
#  FROM Means WHERE PONDNAME='",Lake,"' AND ",ColumnName," IS NOT NULL
# ORDER BY PONDNAME,SAMPLEDATE,SAMPLEDEPTH",sep="")
# data = sqldf(Sql)



#   
# # Perform t-test
# t_test_result <- t.test(data[[ColumnName]] ~ Season, data = data)
# 
# # Print the result
# print(t_test_result)


# 
#       # Perform ANOVA
# anova_result <- aov(DO_Mean ~ Season, data = data)
# 
# # Print the summary of ANOVA
# summary(anova_result)

```

```{r, label="Create site visit plots",results='asis'}
# Loop through each row
Lakes = Visits %>% distinct(PONDNAME)
for (i in 1:nrow(Lakes)) {
  Lake = Lakes[i,"PONDNAME"]
  GetReportSectionForLake(Lake)
}
#GetReportSectionForLake('YUCH-004')

```

# QC Checks on the data

The first step is to check the data quality. 







## QC-`r QCNum`: Water profiles data collection time span

This check shows the timespan in years of water profile data data collection.

```{r}
Sql = "SELECT Min(Year(SAMPLEDATE)) As Earliest
,Max(Year(SAMPLEDATE)) As Latest
,Max(Year(SAMPLEDATE)) - Min(Year(SAMPLEDATE)) As [Years]
FROM tblWaterProfiles"
knitr::kable(dbGetQuery(connection,Sql),caption=paste("Table QC-",QCNum,": Visits history for continuous water quality lakes.",sep=""))
QCNum = QCNum + 1
```

## QC-`r QCNum`: Water profiles data collection time span by Park

This check shows the timespan in years of water profile data data collection by Park.

```{r}
Sql = "SELECT tblPonds.PARK
, MIN(YEAR(tblWaterProfiles.SAMPLEDATE)) AS Earliest
, MAX(YEAR(tblWaterProfiles.SAMPLEDATE)) AS Latest
, MAX(YEAR(tblWaterProfiles.SAMPLEDATE)) - MIN(YEAR(tblWaterProfiles.SAMPLEDATE)) AS Years
FROM tblWaterProfiles 
INNER JOIN tblPonds ON tblWaterProfiles.PONDNAME = tblPonds.PONDNAME
GROUP BY tblPonds.PARK
ORDER BY Years DESC"
knitr::kable(dbGetQuery(connection,Sql),caption=paste("Table QC-",QCNum,": Visits history for continuous water quality lakes by Park.",sep=""))
QCNum = QCNum + 1
```


## QC-`r QCNum`: Number Of Continuous Water Profile Visits By Lake (Which lakes have the longest history?)

This check shows the total number of visits, number of observations and the visitation time span for water profile data collection efforts.

```{r}
Sql = "SELECT * FROM QC_tblWaterProfiles_VisitationHistoryByLake ORDER BY Visits DESC"
knitr::kable(dbGetQuery(connection,Sql),caption=paste("Table QC-",QCNum,": Visits history for continuous water quality lakes.",sep=""))
QCNum = QCNum + 1
```

## QC-`r QCNum`: Minimum and maximum lake depths

Check the minimum and maximum lake depths for anomalous values

```{r}
Sql = "SELECT * FROM QC_tblWaterProfiles_MinMaxDepth"
knitr::kable(dbGetQuery(connection,Sql),caption=paste("Table QC-",QCNum,": Visits history for continuous water quality lakes.",sep=""))
QCNum = QCNum + 1
```

## QC-`r QCNum`: Lake depth intervals

Check the minimum and maximum lake depths for anomalous values

```{r,echo=FALSE,fig.cap=paste("\n\nFigure ",FigureCounter,". Histogram of sampling depths.",sep="")}
Depths = Means %>% select(SAMPLEDEPTH)

# Plot the distribution of depths
ggplot(Depths) + 
  geom_histogram(aes(x=SAMPLEDEPTH),binwidth=1,na.rm=TRUE) + 
  ylab(paste("Count")) + 
  xlab("Depth (m)") + 
  scale_x_continuous(breaks = seq(0.5, 10, by = 0.5)) +
  theme_minimal() +
  ggtitle(paste("Distribution of sampling depths",sep="")) 

FigureCounter = FigureCounter + 1  
QCNum = QCNum + 1
```

## QC-`r QCNum`: Checks on the 0.5m depth dataset

This section extracts only the data at the 0.5m depth since this is what Amy is most interested in. We're looking for anomalous values, outliers or strange patterns that might indicate data quality issues.

```{r,results='asis'}
Sql = "SELECT * FROM Means WHERE SAMPLEDEPTH = 0.5"
Point5Data = sqldf(Sql)

# Loop through the column names
for (ColumnName in colnames(Point5Data)) {
  # Isolate the parameters (they have '_Mean' suffixes)
  if(grepl("_Mean",ColumnName) == TRUE){
  
    # Get a more readable parameter name by removing the '_Mean' suffix and output it as a header
    ParameterName = gsub("_Mean","",ColumnName)
    cat("### ",ParameterName,"  \n\n")
    
    # Plot a histogram showing the distribution of the binned parameter data
    cat("#### Distribution of values  \n\n")
    # Generate some reasonable bin widths for plotting a histogram of the parameter data
    Binwidth <- (max(Point5Data[ColumnName],na.rm=TRUE) - min(Point5Data[ColumnName],na.rm=TRUE)) / ceiling(log2(nrow(Point5Data)) + 1)
    print(ggplot(Point5Data) + 
      geom_histogram(aes_string(x=ColumnName),binwidth=Binwidth,na.rm=TRUE) + 
      ylab(paste("Count (",nrow(Point5Data)," non-null rows)",sep="")) + 
      theme_minimal()) +
      ggtitle(paste("Distribution Of Values: ",ParameterName,sep="")) 
    cat("\n\nFigure ",FigureCounter,"Histogram of ",ParameterName,". Bin width = ",Binwidth,".  \n\n")
    FigureCounter = FigureCounter + 1
    
    # Calculate and summarize in a table basic statistics about the chemical component
    cat("#### Basic statistics  \n\n")
    cat("\n\nTable ",TableCounter,". Basic statistics, ",ParameterName,".  \n\n")
    print(knitr::kable(t(describe(Point5Data[,ColumnName]))))
    cat("  \n\n")
    TableCounter = TableCounter + 1
    
    
    # Plot a histogram showing the distribution of the binned parameter data by season
    cat("#### Distribution of values  \n\n")
    # Generate some reasonable bin widths for plotting a histogram of the parameter data
    Binwidth <- (max(Point5Data[ColumnName],na.rm=TRUE) - min(Point5Data[ColumnName],na.rm=TRUE)) / ceiling(log2(nrow(Point5Data)) + 1)
    print(ggplot(Point5Data) + 
      geom_histogram(aes_string(x=ColumnName),binwidth=Binwidth,na.rm=TRUE) + 
      ylab(paste("Count (",nrow(Point5Data)," non-null rows)",sep="")) + 
      facet_wrap(vars(Season)) +
      theme_minimal()) +
      ggtitle(paste("Distribution Of Values by Park: ",ParameterName,sep="")) 
    cat("\n\nFigure ",FigureCounter,"Histogram of ",ParameterName," by season. Bin width = ",Binwidth,".  \n\n")
    FigureCounter = FigureCounter + 1
    
    # Calculate and summarize in a table basic statistics about the chemical component
    cat("#### Basic statistics, Spring  \n\n")
    cat("\n\nTable ",TableCounter,". Basic statistics, ",ParameterName,".  \n\n")
    SpringData = Point5Data %>% filter(Season=="Spring") %>% select(!!sym(ColumnName),Season)
    print(knitr::kable(t(describe(SpringData))))
    cat("  \n\n")
    TableCounter = TableCounter + 1
    
    # Calculate and summarize in a table basic statistics about the chemical component
    cat("#### Basic statistics, Fall  \n\n")
    cat("\n\nTable ",TableCounter,". Basic statistics, ",ParameterName,".  \n\n")
    FallData = Point5Data %>% filter(Season=="Fall") %>% select(!!sym(ColumnName),Season)
    print(knitr::kable(t(describe(FallData))))
    cat("  \n\n")
    TableCounter = TableCounter + 1
  }
}

QCNum = QCNum + 1
```

## QC-`r QCNum`: Field comments

Field comments often reveal and/or document data defects. There are too many to show here so I put them in Appendix A. 



## QC Checks Summary

A partial compilation of the 500+ field comments from Appendix A that may reveal defects (tblWaterProfiles is not set up for data quality flagging, if anything needs to be fixed then now is the time):

pH readings may be unreliable.  
Be wary of the DO measurements... looks like the DO didn’t recover from the deeper measurements of the previous profile.  
no depth sensor to verify depths.  
DO and Temp may still be stabilizing.  
DO had probably not yet stabilized.
DO looks high; probably had not yet stabilized.
DO may be drifting down; pH readings unreliable.
DO may not yet have stabilized. pH appears to be drifting upwards.
DO measurement is correct; sonde not on bottom during measurement.
DO measurement is suspect.
DO measurements seem implausible; may be down in the muck.
The DO appears anomalous
Drifting.
Sonde may be on bottom.  
Sonde may have been in the muck
There may be scope on the line; depth sensor recorded X.Xm.
Lake very shallow; only surface sample possible.
Looks like pH is oscillating.
May have been collected before DO probe had stabilized. pH readings may be unreliable
Measurement appears to have been collected before probes had stabilized.
TAKEN FROM INSIDE UPWELLINGS
Windy


# Apendix A: Field data comments

The comments below may be useful in deciding if records need repair or exclusion from analysis.

```{r}
knitr::kable(dbGetQuery(connection,"SELECT Distinct PROFILE_COMMENTS FROM tblWaterProfiles WHERE NOT PROFILE_COMMENTS IS NULL And NOT PROFILE_COMMENTS = '' ORDER BY PROFILE_COMMENTS"))
```

# Appendix B: Correspondence regarding the analysis

## Original request from Amy

I am doing a science for lunch on November 4. I would like to do an analysis of the ProDss data at the 0.5 m depth across all the continuous monitoring lakes in CAKN and ARCN. I would like an export of all the data separated into two different tables: one for spring and another for fall.

I would then like to run a regression on each of these different parameters. I am most interested in temperature, spcond and pH. At this time I don't want to try to tackle all the other depths.

Nic- can you create an export of this once the 2024 data are in the database? I think this should be fairly simple. 

Scott- if Nick can get this data would you be able to batch process this in R?

Let me know if you have any questions. If neither of you can help, I will work with James, I have been gifted a few hours of his time to compensate the lack of having an employee to help with this. I appreciate your help; I recognize that my lack of an employee is adding additional burden to you both.

I have been tasked by the region to present these data in preparation for a climate change workshop in February. I wasn't given much notice for this.

Thanks
Amy

## Nick's response

I'm working on processing the 2024 KorDSS continuous profile data.

Last week I finished Python code to convert the Trimble samples feature class (created with Pathfinder by processed by Joel) to a format (from the older Positions software) that could be processed and the imported into the database. I imported the 100 NOAT sample events into table 'tblWaterSamples'.

NOTE: The only reason this processing is important is because in the database the water sample events must be uploaded before the profiles.

However, the spring and fall water sample events (retrievals and deployments), in Trimble, from the continuous data have not been processed yet by Pathfinder. I've sent these to Joel for processing with Pathfinder. He says that he will work on this today. If necessary, I'll figure out how to process them myself with my own installed version of Pathfinder.

The processing of the KorDSS profile data is fast with the VBA utility tool; I ran this data through this utility this morning and will now look for anomalies in the calculated depths. If necessary, depending on how complicated you'd like to have the data exported, this profile data could be appended to any data export; if Joel doesn't get me his processing soon enough (which I think he will).

Nick

## My clarification

Hey Nick. I got Amy's request and your response. I am not up at all on ProDSS which is, I think, the YSI? I see logger deployments and sites, but the continuous data tables are all empty. I think I need an orientation.
 
Those empty continuous tables are not used; that data is in aquarius. Amy is talking about the water profile data 'tblWaterProfiles' related to the water samples table 'tblWaterSamples'. Each event is either continuous or not depending on the value set in the 'tblEvents' column 'CONTMONVISIT' (which indicates instrument deployment or retrieveal). Also, the continuous sites, deployments, and loggers tables aren't related to her requirements.
 
Yes, the ProDSS is YSI.
 
KorDSS is the software that is used to set up the sensors and export the data.

## Amy's clarification on which sites to analyze

The continuous monitoring deploys occur in either May/June depending on spring conditions or September during the fall. This task may be slightly trickier than I thought, the user has to note that it was a continuous monitoring event in the database so some of these events might be missed due to data entry errors.

These are the continuous monitoring sites:
BELA-088
BELA-108
DENA-023
DENA-018
KOVA-085
KOVA-086
NOAT-052
NOAT-107
WRST-005
WRST-016
YUCH-004
YUCH-005
YUCH-082
YUCH-083

There are a handful of other lakes that we tried monitoring continuously but we had deployment problems so we changed sites. These sites all have reasonable long deployment records.

## Amy 2024-10-31

Looks like you have grouped all the continuous data into a single analysis, I am looking for lake by lake comparisons.  These results are interesting, I am more interested in changes over time? Perhaps we do this same analysis by lake?

I'm guessing each of these scripts gives skewness and kurtosis automatically, some water chemistry is notoriously skewed, especially in Alaska, where concentrations are often near zero (no -values are possible and we have detection limits). There are two ways to get around this: non-parametric statistics or transforming our data. pH is already normally distributed because it is the -log of the hydrogen ion concentration.

The last time I did an analysis we used thiel sen slopes, the limitation to these analyses is that all measurements have to be in the same direction or you get insignificant results. I would like to try more a conventional parametric approach this time.  To do so we would need to transform several parameters (SPcond and all the other stuff like TN, TP, Ca....). 

I am not sure how much of this you are interested in pursuing, ultimately what I would like to do is a time-series analysis on each lake  the fall data and a separate analysis using the spring data. I think thus far you are just looking at the four-basic parameters from the profiles. At some point I would like to expand this to include the other constituents (TN, TP...)

Thanks for doing this it is very helpful

Amy

```{r, echo=FALSE,warning=FALSE,fig.cap=paste("Figure ",FigureCounter,". Correlation matrix among parameters.",sep="")}
# Analysis

## Correlation and regression of each parameter against the others


# Plot each parameter against the others 
# ggpairs(MeansDataset, 
#         upper = list(continuous = wrap("cor", size = 3),na.rm=TRUE),
#         lower = list(continuous = wrap("smooth", method = "lm", se = TRUE,size=0.5,color="gray"),na.rm=TRUE)) +
#   theme_minimal()
# FigureCounter = FigureCounter + 1

# Do more or less the same thing using R's pairs() function, but without regression lines and correlation coefficients
# Scatter everything against everything else
#pairs(MeansDataset)

```

```{r, echo=FALSE,results='asis'}
## Seasonality of parameters
# for (Parameter in c("Temperature_Mean","pH_Mean","DO_Mean","SpCond_Mean")){
#   cat("### ",Parameter,"  \n\n")
#   print(
#     ggplot(Point5Data) +
#       #geom_point(aes_string(x='Season',y=Parameter)) +
#       geom_boxplot(aes_string(x='Season',y=Parameter)) +
#       theme_minimal()
#   )
# }



```



