---
title: ""
author: ""
date: ""
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE, echo=TRUE}

# Global NA to blank
options(knitr.kable.NA = '')
knitr::opts_chunk$set(echo = FALSE)

# Libraries
library(sqldf)
library(tidyverse)
library(odbc)
library(leaflet)
library(psych)

```

# DOC vs TN Analysis

Determine if there is a predictable relationship between DOC and DON that can be used to back-generate DON for lakes where it was omitted during data collection

Total Nitrogen (TN): This is the sum of all forms of nitrogen in the water, including organic nitrogen (like proteins and amino acids), inorganic nitrogen (like ammonia, nitrite, and nitrate), and dissolved organic nitrogen (DON).

Total Dissolved Nitrogen (TDN): This includes all forms of nitrogen that are dissolved in the water, excluding particulate nitrogen. It encompasses inorganic nitrogen (ammonia, nitrite, nitrate) and dissolved organic nitrogen (DON).

Inorganic Nitrogen: This refers to nitrogen in its inorganic forms, such as ammonia (NH₃), nitrite (NO₂⁻), and nitrate (NO₃⁻). These forms are typically more bioavailable and can be readily used by aquatic organisms.

# Methods

The water chemistry data are contained in the tblWaterChemistry database table. Water chemistry parameters generally are collected as replicates (A/B/C, etc.) that must be averaged together by lake and sampling date. These calculations are done by the database through a query called Dataset_DOC_DON. The summaries and analyses presented in this report are based on the results of the Dataset_DOC_DON query. 

```{r, echo=TRUE}
# Get the data
# The database has a query called Dataset_DOC_DON which calculates a statistics over water sample replicates for DOC/DON by lake and sample date. 

# Database connection 
connection = dbConnect(odbc(),Driver = "Sql Server",Server = "inpyugamsvm01\\nuna", Database = "AK_ShallowLakes")

# Query the Dataset_DOC_DON database view
Sql = "SELECT d.PONDNAME
, d.PARK
, CONVERT(Date, d.SAMPLEDATE) AS SAMPLEDATE

, d.DOC_Mean
, d.DOC_Min
, d.DOC_Max
, d.DOC_SD
, d.DOC_NumberOfNonNullSamples
, d.[DOC_SampleReplicates (A/B/C)]

, d.TDN_Mean, d.TDN_Min
, d.TDN_Max, d.TDN_SD
, d.TDN_n
, d.TDN_NumberOfNonNullSamples
, d.[TDN_SampleReplicates (A/B/C)]

, d.TN_Mean
, d.TN_Min
, d.TN_Max
, d.TN_SD
, d.TN_n
, d.TN_NumberOfNonNullSamples
, d.[TN_SampleReplicates (A/B/C)]

, d.DON_Mean
, d.DON_Min
, d.DON_Max
, d.DON_SD
, d.DON_n

, case when p.LATITUDE_WGS84 is NULL then p.LATITUDE_NAD83 end AS Lat
, case when p.LONGITUDE_WGS84 is NULL then p.LONGITUDE_NAD83 end AS Lon

FROM Dataset_DOC_DON AS d 
INNER JOIN tblPonds AS p ON d.PONDNAME = p.PONDNAME
ORDER BY d.PONDNAME, SAMPLEDATE"

# Make a data frame
Data = dbGetQuery(connection,Sql)

# Make sampling dates real dates instead of strings
Data$SAMPLEDATE = as.Date(Data$SAMPLEDATE)
# view(Data)

# Get the number of records missing DON
NumberOfRecordsWithDOCButNoDON = sqldf("SELECT Count(*) FROM Data WHERE DOC_Mean IS NOT NULL And (TDN_Mean IS NULL Or TDN_Mean = '')")

```



There are `r NumberOfRecordsWithDOCButNoDON` records with a recorded DON but no TDN value.


# Step 1: QC the chemistry data

## QC the parameters

```{r,echo=FALSE,results='asis'}
# Functions

# Get a report section for a chemistry component name - use only the component name, not trailing stuff like '_MG_L', etc
# Example: GetChemistryComponentQCReportSection('DOC')
GetChemistryComponentQCReportSection = function(ChemistryComponentName){
  
  #ChemistryComponentName='TDN'
  
  # Output a section header
  cat("### ",ChemistryComponentName,"  \n\n")
  
  # Build a column name for the parameter. Example: DOC_Mean
  Parameter = paste(ChemistryComponentName,"_Mean",sep="")

  # Boxplot of the parameter
  cat("**Box Plot**  \n\n")
  Sql = paste("SELECT PARK,",Parameter," FROM Data WHERE ",Parameter," IS NOT NULL",sep="")
  ChemData = sqldf(Sql)
  cat(nrow(ChemData)," non-null rows.  \n\n")
  
  # Box plot of the mean chemical parameter
  print(ggplot(ChemData) + 
          geom_boxplot(aes_string(y=Parameter),na.rm=TRUE) + 
          theme_minimal())
  cat("  \n\n")
  cat(paste("\n\nFigure X. Box plot of ",Parameter,".  \n\n",sep=""))
  
  # Box plot of the parameter faceted by Park
  print(ggplot(ChemData) + 
          geom_boxplot(aes_string(y=Parameter),na.rm=TRUE) + 
          facet_wrap(vars(PARK)) + 
          theme_minimal())
  cat("  \n\n")
  cat(paste("\n\nFigure X. Box plot of ",Parameter," by park.  \n\n",sep=""))
  cat("  \n\n")
  
  
  
  # Plot a curve of the parameter to show normality
  cat("**Histogram of values**  \n\n")
  cat(nrow(ChemData)," non-null rows.  \n\n")
  # Calculate a reasonable bin width for the histogram below
  Binwidth <- (max(ChemData[Parameter]) - min(ChemData[Parameter])) / ceiling(log2(nrow(ChemData)) + 1)
  # Plot a histogram showing the distribution of the binned chemical data
  print(ggplot(ChemData) + 
          geom_histogram(aes_string(x=Parameter),binwidth=Binwidth,na.rm=TRUE) + 
          ylab(paste("Count (",nrow(ChemData)," non-null rows)",sep="")) + 
          theme_minimal())
  cat(paste("\n\nFigure X. Histogram of ",Parameter,". Bin width = ",Binwidth,".  \n\n",sep=""))
  cat("  \n\n")
  # Same histogram as above but faceted by park
  print(ggplot(ChemData) + 
          geom_histogram(aes_string(x=Parameter),binwidth=Binwidth,na.rm=TRUE) + 
          ylab(paste("Count (",nrow(ChemData)," non-null rows)",sep="")) + 
          facet_wrap(vars(PARK)) +
          theme_minimal())
  cat(paste("\n\nFigure X. Histogram of ",Parameter," by Park. Bin width = ",Binwidth,".  \n\n",sep=""))
  cat("  \n\n")
  
  # Calculate and summarize in a table basic statistics about the chemical component
  cat("**Basic statistics: ",ChemistryComponentName,"**  \n\n")
  stats = sqldf(paste("SELECT ",Parameter," FROM Data WHERE ",Parameter," IS NOT NULL ORDER BY ",Parameter," DESC",sep=""))
  cat("  \n\n")
  print(knitr::kable(t(describe(stats))))
  cat("  \n\n")
  
  
  # Determine what minimum detection levels exist for the parameter
  # Append '_MDL_LEVEL' to the component name: Example DOC_MDL_LEVEL
  MDLColumnName = paste(ChemistryComponentName,"_MDL_LEVEL",sep="")
  cat("**Minimum detection level(s): ",MDLColumnName,"**  \n\n")
  # Get the distinct MDL for the chemical component. These are usually consistent by lab, so lab changes are where these MDLs are likely to change
  Sql = paste("SELECT DISTINCT ",MDLColumnName," As MDL,CHEM_LAB FROM tblWaterChemistry WHERE ",MDLColumnName," IS NOT NULL ORDER BY ",MDLColumnName,sep="")
  # Dump out  table of Lab/MDLs for the chemical component
  MDLs = dbGetQuery(connection,Sql)
  print(knitr::kable(MDLs))
  cat("\n\n")
  #cat(Sql,"  \n\n")

  # Show the number of records, if any, that are below MDL
  cat("**Number of records below minimum detection level**  \n\n")
  Sql = paste("SELECT COUNT(*) AS [n] FROM tblWaterChemistry WHERE (",ChemistryComponentName,"_MG_L < ",ChemistryComponentName,"_MDL_LEVEL)",sep="")
  print(knitr::kable(dbGetQuery(connection,Sql)))
  cat("\n\n")
  #cat(Sql,"  \n\n")

  # Comments that might be important to know
  cat("**Comments from the database for: ",ChemistryComponentName,"**  \n\n")
  cat("The comments below may be useful in determining any records to exclude or modify. These comments may provide insight about any outliers or data anomalies.  \n\n")
  Sql = paste("SELECT DISTINCT ",ChemistryComponentName,"_Notes FROM tblWaterChemistry WHERE (",ChemistryComponentName,"_Notes IS NOT NULL) and LTRIM(RTRIM(",ChemistryComponentName,"_Notes)) <> '' And ISNUMERIC(",ChemistryComponentName,"_Notes) = 0 ORDER BY ",ChemistryComponentName,"_Notes",sep="")
  print(knitr::kable(dbGetQuery(connection,Sql)))
  cat("  \n\n")

}

```

```{r,results='asis'}
GetChemistryComponentQCReportSection('DOC')
GetChemistryComponentQCReportSection('TDN')
GetChemistryComponentQCReportSection('TN')

```

### DON

DON is calculated as TN - TDN. A lot of outliers exist, particularly in YUCH which skew DON far to the right.

```{r}

# Box plot of the mean chemical parameter
ggplot(Data) + 
 geom_boxplot(aes(y=DON_Mean),na.rm=TRUE) + 
 theme_minimal()

# Box plots faceted by park
ggplot(Data) + 
  geom_boxplot(aes(y=DON_Mean),na.rm=TRUE) + 
  facet_wrap(vars(PARK)) +
  theme_minimal()

# Calculate a reasonable bin width for the histogram below
Binwidth <- (max(Data$DON_Mean,na.rm=TRUE) - min(Data$DON_Mean,na.rm=TRUE)) / ceiling(log2(nrow(Data)) + 1)
# Plot a histogram showing the distribution of the binned chemical data
ggplot(Data) + 
  geom_histogram(aes(x=DON_Mean),binwidth=Binwidth,na.rm=TRUE)  + 
  theme_minimal()
  
# Calculate a reasonable bin width for the histogram below
Binwidth <- (max(Data$DON_Mean,na.rm=TRUE) - min(Data$DON_Mean,na.rm=TRUE)) / ceiling(log2(nrow(Data)) + 1)
# Plot a histogram showing the distribution of the binned chemical data
ggplot(Data) + 
  geom_histogram(aes(x=DON_Mean),binwidth=Binwidth,na.rm=TRUE)  + 
  facet_wrap(vars(PARK)) +
  theme_minimal()

  # Calculate and summarize in a table basic statistics about the chemical component
  stats = sqldf(paste("SELECT TN_Mean - TDN_Mean As DON FROM Data ORDER BY DON DESC",sep=""))
  knitr::kable(t(describe(stats)))
  
```


## QC: Dissolved Nitrogen Greater Than Total Nitrogen

```{r, echo=TRUE}
# Get a dataset of records having total dissolved nitrogen > total nitrogen
TDNGTTN = dbGetQuery(connection,"SELECT Park,  [PONDNAME]
      ,[SAMPLEDATE]
      ,[SAMPLENUMBER]
      ,[CHEM_LAB]
      ,[CHEM_METHOD]
      ,[TDN_MG_L]
      ,[TN_MG_L]
      ,[TDN_Notes]
      ,[TN_Notes]
      ,[NOTES]
      ,[COMMENTS]
      ,[DOC_LABDATE]
      ,[LAB_SAMPLE_NUMBER]
      ,[LAB_DATE]
      ,[DELIVERY_DATE]
      ,[THAW_DATE]
      ,[DUPLICATED_CHEM_ANALYSIS]
  FROM [AK_ShallowLakes].[dbo].[QC_tblWaterChemistry_TDN_GreaterThan_TN]")


```

`r nrow(TDNGTTN)` records have total dissolved nitrogen greater than total nitrogen.

Table X .Summary of offending records.

```{r}
Sql = "SELECT        PARK, COUNT(*) AS n
FROM            QC_tblWaterChemistry_TDN_GreaterThan_TN
GROUP BY PARK
ORDER BY n desc"
knitr::kable(dbGetQuery(connection,Sql))

```


```{r, echo=FALSE,include=FALSE}
# I wanted to look at the data too see if there were any spatial explanations for the outliers but it didn't work well
# and many plots were lacking coordinates.
# library(ggspatial)
# ggplot(Data %>% filter(PARK == 'YUCH')) +
#   geom_point(aes(x=Lat,y=Lon,size=DOC_Mean,color=TDN_Mean),na.rm=TRUE) +
#   scale_size_continuous(range = c(0,3)) +
#   geom_spatial_wms('https://basemap.nationalmap.gov:443/arcgis/services/USGSTopo/MapServer/WmsServer', layers = "0") +
#   theme_minimal()
  #facet_wrap(vars(PARK),ncol=1) 
  # +
  # theme(#panel.spacing = unit(4, "cm"),  # Adjust space between plots
  #   #plot.margin = unit(c(1, 1, 1, 1), "cm"),  # Adjust plot margins
  #   aspect.ratio = 1 # Set the aspect ratio ) 

```


## QC Summary

- Neither DOC nor TDN contain records with values below laboratory mean detection levels. TN does contain records below MDL, but it looks like Heidi set some of these values at just below threshold (0.99 where MDL is 1, for example). I think these records can safely be included.

- DOC, TDN and TN have positive skewness (skewing toward larger values) and TDN and TN have high kurtosis (tall skinny distribution) as well. Boxplots show a lot of outliers.
 
- YUCH seems to have more elevated DOC than other parks, and many more outliers.

- NOAT is the only park with low value outliers for DOC.

- Some records had a note: "Carbon Analysis run outside of hold time due to instrumentation problems".

- Many records exist with TDN \> TN and should be investigated

- After calculating DON as TN - TDN I'm not sure this gets us anything since TDN is typically a very large part of DON.


# Step 2: Scatter Plots and Regressions

## Linear Model

### DOC Vs TDN

```{r, echo=TRUE}
# Plot mean DOC per site and sampling date against mean TDN per site and sampling date
ggplot(Data) +
  geom_point(aes(x=DOC_Mean,y=TDN_Mean,color=PARK),na.rm=TRUE) +
  geom_smooth(aes(x=DOC_Mean,y=TDN_Mean),na.rm=TRUE,se =TRUE, method='lm') +
  theme_minimal()

# Fit a linear regression model
model <- lm(Data$DOC_Mean ~ Data$TDN_Mean, data = Data)

# Display the model summary
summary(model)


```

Linear models by Park

```{r}
# Plot mean DOC per site and sampling date against mean TDN per site and sampling date
ggplot(Data) +
  geom_point(aes(x=DOC_Mean,y=TDN_Mean),na.rm=TRUE) +
  geom_smooth(aes(x=DOC_Mean,y=TDN_Mean),na.rm=TRUE,se =TRUE, method='lm') +
  facet_wrap(vars(PARK)) +
  theme_minimal()


```

### DOC Vs TN

```{r, echo=TRUE}
# Plot mean DOC per site and sampling date against mean TN per site and sampling date
ggplot(Data) +
  geom_point(aes(x=DOC_Mean,y=TN_Mean,color=PARK),na.rm=TRUE) +
  geom_smooth(aes(x=DOC_Mean,y=TN_Mean),na.rm=TRUE,se =TRUE, method='lm') +
  theme_minimal()

# Fit a linear regression model
model <- lm(Data$DOC_Mean ~ Data$TN_Mean, data = Data)

# Display the model summary
summary(model)


```

Linear models by Park

```{r}
# Plot mean DOC per site and sampling date against mean TN per site and sampling date
ggplot(Data) +
  geom_point(aes(x=DOC_Mean,y=TN_Mean),na.rm=TRUE) +
  geom_smooth(aes(x=DOC_Mean,y=TN_Mean),na.rm=TRUE,se =TRUE, method='lm') +
  facet_wrap(vars(PARK)) +
  theme_minimal()


```

### DOC Vs DON

DOC vs DON: From the data it seems that TDN is a very large component of TN, tracking it closely, in some cases exceeding it giving negative DON values. I think we need to look at this before concluding anything.

The linear model is significant but R^2 is surprisingly low. 

```{r, echo=TRUE}
# Plot mean DOC per site and sampling date against mean DON (TN-TDN) per site and sampling date
ggplot(Data) +
  geom_point(aes(x=DOC_Mean,y=DON_Mean,color=PARK),na.rm=TRUE) +
  geom_smooth(aes(x=DOC_Mean,y=DON_Mean),na.rm=TRUE,se =TRUE, method='lm') +
  theme_minimal()

# Fit a linear regression model
model <- lm(Data$DOC_Mean ~ Data$DON_Mean, data = Data)

# Display the model summary
summary(model)

# To see the data:
#knitr::kable(Data %>% filter(is.na(DOC_Mean) == FALSE & is.na(DON_Mean) == FALSE) %>% select(DOC_Mean,TN_Mean,TDN_Mean,DON_Mean))

```


Linear models by Park

```{r}
# Plot mean DOC per site and sampling date against mean TDN per site and sampling date
ggplot(Data) +
  geom_point(aes(x=DOC_Mean,y=DON_Mean),na.rm=TRUE) +
  geom_smooth(aes(x=DOC_Mean,y=DON_Mean),na.rm=TRUE,se =TRUE, method='lm') +
  facet_wrap(vars(PARK)) +
  theme_minimal()


```

## Initial Conclusions

-   Given the dataset's potential problems, the R\^2 and P values indicate a significant linear relationship of moderate explanatory power. Dealing with the records having TDN > TN and outlier removal may improve R^2 fit.

-   It looks like YUCH has, if not outliers, more extreme values. Need to investigate.

- After calculating DON as TN - TDN I'm not sure this gets us anything since TDN is typically a very large part of DON.

